Security Metrics for the Android Ecosystem

### 2015-07-27 security_scoring.tex
Corrections on reading through

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/proportioninsecure}
\caption{Proportion of devices running insecure, maybe secure and secure versions of Android.
Table~\ref{tab:andvulns} lists the \daNumVulnsUsed\ vulnerabilities used, the red vertical lines are caused by their discovery and the most important are annotated.}
\label{fig:proportioninsecure}
\end{figure}
By combining data on critical vulnerabilities in Android and the versions of Android running on devices we can determine which vulnerabilities each device was vulnerable to each day.
Then we consider a device is insecure if it is running a vulnerable version of Android and the device has not received an update which might fix it;
it is maybe secure if it is running a vulnerable version but received an update which could have fixed the vulnerability if it contained a backported fix;
and it is secure if it is running a secure version.
This allows us to plot Figure~\ref{fig:proportioninsecure}, initially all devices are \emph{maybe secure} (yellow) since \da\ does not have historical data prior to May 2011.
This means we cannot distinguish between devices which are running a version of Android which is known to be vulnerable from one which may have received a backported fix.
This demonstrates the importance of a longitudinal study: this type of analysis requires years of data.
Once \avovuln{zergRush} was discovered in October 2011 then most devices are recorded as \emph{insecure} (red) as they were exposed to that vulnerability.
The remaining devices were already running a version of Android which fixed the \avovuln{zergRush} vulnerability and are therefore marked as \emph{secure} (green).
From October 2011 until the discovery of \avovuln[APK_duplicate_file]{APK duplicate file} in February 2013 the graph shows progressive improvement as devices are upgraded.
This means more and more devices are marked as \emph{secure} because they are now running a secure version of Android, or marked as \emph{maybe secure} because they received an OS update that did not update to a known-good version of Android but which may still have included a backport of a fix, as the update was made available after the vulnerability was disclosed.
From February 2013 onwards regular discovery of critical vulnerabilities ensures that most devices are exposed to known critical vulnerabilities.
This gives us an average exposure to critical vulnerabilities of Android devices of \daMeanInsecurityPerc\ (this percentage ignores devices classed as \emph{maybe secure}).
This is $1 - f$, the proportion of devices free from critical vulnerabilities, however this is only one component of a more robust score, the FUM score.

@@

However there are a wide variety of scores depending on the source of the device.
There is anecdotal evidence that Google's Nexus devices are better at getting updates than other Android devices because Google makes the original updates and ships them to its devices\footnote{\url{http://www.howtogeek.com/139391/htg-explains-why-android-geeks-buy-nexus-devices/}}.
Table~\ref{tab:sec_summary} shows that this is the case with Nexus devices getting much better scores than non-Nexus devices and Figure~\ref{fig:security_score_summary} shows this holds true over time.

>>>>

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/proportioninsecure}
\caption{Proportion of devices running insecure, maybe secure and secure versions of Android.
Table~\ref{tab:andvulns} lists the \daNumVulnsUsed\ vulnerabilities used, the red vertical lines are caused by their discovery and the most important are annotated.}
\label{fig:proportioninsecure}
\end{figure}
By combining data on critical vulnerabilities in Android and the versions of Android running on devices we can determine which vulnerabilities each device was vulnerable to each day.
Then we consider a device is \emph{insecure} if it is running a vulnerable version of Android and the device has not received an update which might fix it;
it is \emph{maybe secure} if it is running a vulnerable version but received an update which could have fixed the vulnerability if it contained a backported fix;
and it is \emph{secure} if it is running a secure version.
This allows us to plot Figure~\ref{fig:proportioninsecure}, initially all devices are \emph{maybe secure} (yellow) since \da\ does not have historical data prior to May 2011.
This means we cannot distinguish between devices which are running a version of Android which is known to be vulnerable from one which may have received a backported fix.
This demonstrates the importance of a longitudinal study: this type of analysis requires years of data.
Once \avovuln{zergRush} was discovered in October 2011 then most devices are recorded as \emph{insecure} (red) as they were exposed to that vulnerability.
The remaining devices were already running a version of Android which fixed the \avovuln{zergRush} vulnerability and are therefore marked as \emph{secure} (green).
From October 2011 until the discovery of \avovuln[APK_duplicate_file]{APK duplicate file} in February 2013 the graph shows progressive improvement as devices are upgraded.
This means more and more devices are marked as \emph{secure} because they are now running a secure version of Android, or marked as \emph{maybe secure} because they received an OS update that did not update to a known-good version of Android but which may still have included a backport of a fix, as the update was made available after the vulnerability was disclosed.
From February 2013 onwards regular discovery of critical vulnerabilities ensures that most devices are exposed to known critical vulnerabilities.
This gives us an average exposure to critical vulnerabilities of Android devices of \daMeanInsecurityPerc\ (this percentage ignores devices classed as \emph{maybe secure}).
This is $1 - f$, the proportion of devices free from critical vulnerabilities, however this is only one component of a more robust score, the FUM score.

@@

However there are a wide variety of scores depending on the source of the device.
There is anecdotal evidence that Google's Nexus devices are better at getting updates than other Android devices because Google makes the original updates and ships them to its devices\footnote{\url{http://www.howtogeek.com/139391/htg-explains-why-android-geeks-buy-nexus-devices/}}.
Table~\ref{tab:sec_summary} shows that this is the case with Nexus devices getting much better scores than non-Nexus devices.

### 2015-07-27 spsm04-thomas.tex
Corrections on reading through

\begin{abstract}
The security of Android depends on timely delivery of security updates for critical vulnerabilities .
We define the FUM security metric to rank the performance of device manufacturers and network operators, based on their provision of updates and exposure to critical vulnerabilities.
Using a corpus of \daNumOSDevices\ devices we show that there is significant variability in the timely delivery of security updates across different device manufacturers and network operators.
This provides a comparison point for purchasers to determine which device manufacturers and network operators to use.
We find that \daMeanInsecurityPercNominal\ of Android devices are exposed to at least one of \daNumVulnsUsed\ known critical vulnerabilities and assign a FUM security score of \daSecurityScoreNominal\ out of 10 while Nexus devices gets \daSecScoreBestsummaryScoreNominal\ and \daSecScoreBestmanufacturer\ \daSecScoreBestmanufacturerScoreNominal.
On average \daMeanOutstandingVulnerabilitiesNominal\ outstanding vulnerabilities are not fixed on any device and only \daUpdatednessPercNominal\ of devices run the current version of Android.
\end{abstract}

@@

\todo{new opening paragraph}

@@

The Android update process is slow because its development is distributed among many parties and has a long multi-stage pipeline~\cite{HTC2013}.
The security of Android relies on many open source projects, including the Li\-nux kernel, Open\-SSL and Bouncy\-Castle, as well as Google, who build the core platform.
In addition, device manufacturers (we know of \daNumManufacturers) and network operators (\daNumOperators) may make further modifications before devices are shipped to customers.
Understanding this ecosystem is important because device manufacturers have introduced additional vulnerabilities in the past~\cite{Grace2012}.
The cause of the slow update process could be: upstream open source projects releasing patches slowly, Google including them slowly, manufacturers including them slowly, network operators shipping them slowly or customers not installing them.
We present a better understanding of both the ecosystem of Android development and associated vulnerabilities more fully in~\S\ref{sec:android_ecosystem}.

By using data from \da~\cite{Wagner2013} we can examine the versions of Android running on user devices over time and through comparison with data on which versions of Android were vulnerable to different critical vulnerabilities~\cite{androidvulnerabilities.org} we can plot the exposure of Android devices to known critical vulnerabilities.
Figure~\ref{fig:proportioninsecure} uses data from \da\ (\S\ref{sec:da}) and the AndroidVulnerabilities.org (\avo) vulnerabilities database (\S\ref{sec:avo}) and shows that Android devices are often exposed to known security vulnerabilities (\S\ref{sec:androidvulnerability}).
One solution to this problem is regulation, and indeed there is ongoing legal action to force network operators to ship updates for security vulnerabilities~\cite{Soghoian2013}.\todolater{Check on the status of this legal action}
Many smartphones are sold on 12--24 month contracts, and yet our data shows many devices do not receive many security updates, with an overall average of \daUpdatesPerYearNominal\ per year. 
In contrast, Windows XP could be purchased for a one-off payment in October 2001 and received security updates monthly until April 2014.
An alternative to regulation is access to publicly available data.
Comparative data, which we provide in this paper, and will continue to publish on our companion website, provides an incentive to both device manufacturers and network operators to provide updates.

In summary, the contributions of this paper are:
\begin{itemize}
 \item We quantify the Android update process, providing concrete numbers on the flow of updates and their latency~(\S\ref{sec:android_ecosystem}).
 \item We determine that the main update bottleneck lies with manufacturers rather than operators or users~(\S\ref{sec:bottleneck}).
 \item We propose the FUM scoring metric to evaluate the security of different instances of a platform~(\S\ref{sec:security_scoring:method}).
 \item We measure the security of Android against our scoring metric and compare different device manufacturers, device models and network operators to allow device purchasers to differentiate between them based on security~(\S\ref{sec:security_scoring:results}).
\end{itemize}

@@

The best place to prevent attacks is at runtime, since all three attack vectors can be prevented at this level. 
Unfortunately, as Figure~\ref{fig:proportioninsecure} shows, the sandbox for Android apps is weak, since on most devices there are known critical vulnerabilities in it.

\section{Data}
We use two sources of data to measure the security of Android: (1) information on the distribution of Android versions over time and (2) information on the critical vulnerabilities found to affect Android.
These two datasets can then be combined to determine the proportion of devices at risk of attack from specific vulnerabilities.

@@

Some critical vulnerabilities are not traditional kernel vulnerabilities, but are vulnerabilities in \emph{installation} in our threat model.
For example improper verification of signatures at installation time was discovered in February 2013~\cite{Forristal2013} and meant that applications could pretend to be signed with system keys and hence be granted system privileges.
On versions of Android below 4.1 malware could then use known system-to-root escalation mechanisms.
Regardless of version, this exposed an increased attack area and would also provide the ability for malware to control all user internet traffic (via VPNs), brick the phone, remove and install apps, steal user credentials and read the screen.
The different categories in which the vulnerabilities fall are shown in Table~\ref{tab:andvulns}, `signature' vulnerabilities are \emph{installation}, while kernel and system vulnerabilities are used with \emph{installation, dynamic code loading} and \emph{injection}.
\avoTabAndVulns

@@

\subsection{Device Analyzer data}\label{sec:da}

In this analysis we use historical data collected by the \da\ project~\cite{Wagner2013}.
\da\ collects data\footnote{\url{https://deviceanalyzer.cl.cam.ac.uk/collected.html}} from study participants who install the Android app from Google Play.
Most study participants allow researchers a\-round the world to access a subset of their device data, including the data presented here.

@@

\da\ has collected data from \daNumDevices\ devices with a total of \daDeviceDays\footnote{Here we are only counting devices and days for which we have valid OS version data.} device days.
The majority of devices only contribute data for a short period of time, however \daMonthsDevices\ devices have contributed data for more than \daMonths~months.

@@

\section{Comparison with other data}\label{sec:representative}
Here we provide evidence that the \da\ data~\cite{Wagner2013} is representative for our purposes by comparing it with three other sources of data.
The \da\ data is collected from devices across the world where users have voluntarily installed the app.
We have obtained comparable data on 5\,290 devices from a multinational FTSE 100 company's mobile device management database, which includes company and employee owned devices, and from 5\,170\,000 matching User-Agent headers on all HTTP traffic for 30\% of Rwanda for a week.
\todo{citation for Rwanda dataset}
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/dists}
\caption{Comparison between FTSE, User-Agent and the corresponding \da\ data, error bars indicate 95\% confidence intervals.}
\label{fig:dists}
\end{figure}
We used the data from the FTSE 100 company for a week in April 2015 and the User-Agent data was collected in February 2015. %between 2015-01-31 and 2015-02-12.% FTSE 2015-04-27
Figure~\ref{fig:dists} shows the proportion of devices running each Android OS version in the two comparison data sets and the comparable periods from \da.
The general pattern this shows is that in the FTSE data newer versions are more popular than in the \da\ data and that in the Rwanda data old versions are more popular than in the \da\ data.
Therefore the \da\ data on OS versions is bounded by these two data sets.
\todolater{Can we quantify this?}

>>>>

\begin{abstract}
The security of Android depends on timely delivery of security updates for critical vulnerabilities .
We define the FUM security metric to rank the performance of device manufacturers and network operators, based on their provision of updates and exposure to critical vulnerabilities.
Using a corpus of \daNumOSDevices\ devices we show that there is significant variability in the timely delivery of security updates across different device manufacturers and network operators.
This provides a comparison point for purchasers to determine which device manufacturers and network operators to use.
We find that on average \daMeanInsecurityPercNominal\ of Android devices are exposed to at least one of \daNumVulnsUsed\ known critical vulnerabilities and assign a FUM security score of \daSecurityScoreNominal\ out of 10 while Nexus devices gets \daSecScoreBestsummaryScoreNominal\ and \daSecScoreBestmanufacturer\ \daSecScoreBestmanufacturerScoreNominal.
On average \daMeanOutstandingVulnerabilitiesNominal\ outstanding vulnerabilities are not fixed on any device and on average only \daUpdatednessPercNominal\ of devices run the current version of Android.
\end{abstract}

@@

To evaluate the security of a system we need to be able to measure it.
Only then can customers and regulators behave appropriately and without it then they are stuck with a lemons market.

@@

The Android update process is slow because its development is distributed among many parties and has a long multi-stage pipeline~\cite{HTC2013}.
The security of Android relies on many open source projects, including the Li\-nux kernel, Open\-SSL and Bouncy\-Castle, as well as Google, who build the core platform.
In addition, device manufacturers (we know of \daNumManufacturers) and network operators (\daNumOperators) may make further modifications before devices are shipped to customers.
Understanding this ecosystem is important because device manufacturers have introduced additional vulnerabilities in the past~\cite{Grace2012}.
The cause of the slow update process could be: upstream open source projects releasing patches slowly, Google including them slowly, manufacturers including them slowly, network operators shipping them slowly or customers not installing them.
We present a better understanding of both the ecosystem of Android development and associated vulnerabilities more fully in~\S\ref{sec:android_ecosystem} and the update bottleneck in \S\ref{sec:bottleneck}.

By using data from \da~\cite{Wagner2013} we can examine the versions of Android running on user devices over time and through comparison with data on which versions of Android were vulnerable to different critical vulnerabilities~\cite{androidvulnerabilities.org} we can measure the exposure of Android devices to known critical vulnerabilities.
One solution to this problem is regulation, and indeed there is ongoing legal action to force network operators to ship updates for security vulnerabilities~\cite{Soghoian2013}.\todolater{Check on the status of this legal action}
Many smartphones are sold on 12--24 month contracts, and yet our data shows many devices do not receive many security updates, with an overall average of \daUpdatesPerYearNominal\ updates per year. 
In contrast, Windows XP could be purchased for a one-off payment in October 2001 and received security updates monthly until April 2014.
An alternative to regulation is access to publicly available data.
Comparative data, which we provide in this paper, and will continue to publish on our companion website, provides an incentive to both device manufacturers and network operators to provide updates.

In summary, the contributions of this paper are:
\begin{itemize}
 \item We quantify the Android update process, providing concrete numbers on the flow of updates and their latency~(\S\ref{sec:android_ecosystem}).
 \item We propose the FUM scoring metric to evaluate the security of different instances of a platform~(\S\ref{sec:security_scoring:method}).
 \item We measure the security of Android against our scoring metric and compare different device manufacturers, device models and network operators to allow device purchasers to differentiate between them based on security~(\S\ref{sec:security_scoring:results}).
 \item We determine that the main update bottleneck lies with manufacturers rather than operators or users~(\S\ref{sec:bottleneck}).
\end{itemize}

@@

The best place to prevent attacks is at runtime, since all three attack vectors can be prevented at this level. 
Unfortunately, the sandbox for Android apps is weak, since on most devices there are known critical vulnerabilities in it.

\section{Data}
We use two sources of data to measure the security of Android: (1) information on the critical vulnerabilities found to affect Android and (2) information on the distribution of Android versions over time.
These two datasets can then be combined to determine the proportion of devices at risk of attack from specific vulnerabilities.

@@

Some critical vulnerabilities are not traditional kernel vulnerabilities, but are vulnerabilities in \emph{installation} in our threat model.
For example improper verification of signatures at installation time was discovered in February 2013~\cite{Forristal2013} and meant that applications could pretend to be signed with system keys and hence be granted system privileges.
On versions of Android below 4.1 malware could then use known system-to-root escalation mechanisms.
Regardless of version, this exposed an increased attack area and would also provide the ability for malware to control all user internet traffic (via VPNs), brick the phone, remove and install apps, steal user credentials and read the screen.
The different categories in which the vulnerabilities fall are shown in Table~\ref{tab:andvulns}, `signature' vulnerabilities are \emph{installation}, while `kernel' and `system' vulnerabilities are used with \emph{installation, dynamic code loading} and \emph{injection}.
\avoTabAndVulns

@@

\subsection{Device Analyzer data}\label{sec:da}
We use historical data collected by the \da\ project~\cite{Wagner2013}.
\da\ collects data\footnote{\url{https://deviceanalyzer.cl.cam.ac.uk/collected.html}} from study participants who install the Android app from Google Play.
Most study participants allow researchers a\-round the world to access a subset of their device data, including the data presented here.

@@

\da\ has collected data from \daNumDevices\ devices with a total of \daDeviceDays device days.
The majority of devices only contribute data for a short period of time, however \daMonthsDevices\ devices have contributed data for more than \daMonths~months.

@@

\section{Comparison with other data}\label{sec:representative}
We compare the Device Analyzer data with three other sources of data and show it is representative for our purposes.
We have obtained comparable data on 5\,290 devices from a multinational FTSE 100 company's mobile device management database, which includes company and employee owned devices, and from 5\,170\,000 matching User-Agent headers on all HTTP traffic for 30\% of Rwanda for a week.
\todo{citation for Rwanda dataset}
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/dists}
\caption{Comparison between FTSE, User-Agent and the corresponding \da\ data, error bars indicate 95\% confidence intervals.}
\label{fig:dists}
\end{figure}
We used the data from the FTSE 100 company for a week in April 2015 and the User-Agent data was collected in February 2015. %between 2015-01-31 and 2015-02-12.% FTSE 2015-04-27
Figure~\ref{fig:dists} shows the proportion of devices running each Android OS version in the two comparison data sets and the comparable periods from \da.
The general pattern this shows is that in the FTSE data newer versions are more popular than in the \da\ data and that in the Rwanda data old versions are more popular.
Therefore the \da\ data on OS versions is bounded by these two data sets.
\todolater{Can we quantify this?}
